biclaR: Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools

### 2023-11-22 paper/PaperCEUS/PaperCEUS.tex
paper CEUS setup

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.21 of 2022/01/12
%
\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.

@@

\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{hyperref}
\hypersetup{colorlinks = TRUE,  urlcolor = blue, linkcolor = blue, citecolor = blue}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
\usepackage{hyperref}
\usepackage{color}
\renewcommand\UrlFont{\color{blue}\rmfamily}

@@

\title{biclaR: Estimating the socio-environmental impacts of car
substitution by bicycle and public transit using open tools}
%
\titlerunning{biclaR: SE impacts of car substitution by bicycle and
public transit}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Rosa Félix\inst{1}\orcidID{0000-0002-5642-6006} \and Filipe
Moura\inst{1}\orcidID{0000-0001-7749-8490} \and Robin
Lovelace\inst{2}\orcidID{0000-0001-5679-6536}}

\authorrunning{R. Félix et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%

\institute{CERIS - Instituto Superior Técnico, University of Lisbon. Av
Rovisco Pais 1049-001 Lisboa, Portugal\\
\email{\href{mailto:rosamfelix@tecnico.ulisboa.pt}{\nolinkurl{rosamfelix@tecnico.ulisboa.pt}}}\\ \and Institute
for Transport Studies, University of Leeds. 34-40 University Rd, Leeds
LS2 9JT, UK}

\maketitle              % typeset the header of the contribution
%
\begin{abstract}
A high proportion of car trips can be replaced by a combination of
public transit and cycling for the first-and-last mile. This paper
estimates the potential for cycling combined with public transit (PT) as
a substitute for car trips in the Lisbon metropolitan area and assesses
its socio-environmental impacts using open data and open source tools. A
decision support tool that facilitates the design and development of a
metropolitan cycling network was developed (\emph{biclaR}). The social
and environmental impacts were assessed using the \emph{HEAT for
Cycling} and the \emph{HEAT as a Service} tools. The impacts of shifting
car trips to PT were also estimated and monetized. The results indicate
that 10\% of trips could be made by bicycle + PT combination. Shifting
to cycling for the first-and-last mile stages can reduce annual
CO\textsubscript{2}eq emissions from 6,000 tons/day, with benefits over
10 years of at least €230 million. For the PT leg, the transfer from car
avoids of up to 20,500 tons of CO\textsubscript{2}eq emissions per year.
This evidence can support policymakers to prioritize interventions that
reduce the reliance on private motor vehicles.

\keywords{Active transport \and Intermodality \and First and last
mile \and Health economic assessment \and Environmental
impacts \and Open data and methods}

\end{abstract}

@@

Combining public transportation (PT) and cycling for the first and last
mile in metropolitan areas can significantly replace private car trips
{[}1, 2{]}. This approach requires interventions and programs to make
bicycling more appealing, and the resulting public investments can have
significant social and environmental benefits.

According to the latest mobility survey conducted in 2018 {[}3{]}, the
LMA registered a total of 5.3 million daily trips, with only 0.5\% by
bicycle. Car modal share was 58.4\%, while PT accounted for 15.5\%. The
number of intra-municipal trips --- with origin and destination in the
same municipality --- amounts to 3.5 million trips. This exceeds the
number of inter-municipal trips (1.8 million trips), involving travel
between different municipalities. Cars and public transport are the most
used modes for intercity trips, with cars being the predominant choice
for all journeys.

To achieve the cycling targets set by the Portuguese national cycling
strategy for 2025 and 2030 (4\% and 10\%, respectively) {[}4{]}, the
Lisbon's Metropolitan Department of Transport introduced
\emph{biclaR}\footnote{See
  \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}.}, a
decision support tool that facilitates the design and development of a
metropolitan cycling network {[}5{]}.

\emph{biclaR} builds on the Propensity to Cycle Tool\footnote{See
  \href{https://www.pct.bike/}{pct.bike}.} (PCT), a web application and
research project funded by the UK's Department for Transport in 2015
which launched nationally in 2017 as part of the government's Cycling
and Walking Investment Strategy. The PCT initially used only
origin-destination data for commuting trips as the basis of estimates of
cycling potential at zone, route and route network levels {[}6{]}. The
PCT has been extended to include cycling potential for travel to school
in England {[}7{]} and other trip types in other countries.\footnote{See
  \href{https://www.npt.scot}{npt.scot} and
  \href{https://cruse.bike}{cruse.bike} for examples of the PCT in
  Scotland and Ireland that include estimates of cycling for other
  purposes.} However, to the best of our knowledge, this is the first
time that the method has been integrated with public transport data
using multi-modal routing to estimate the potential and benefits of
multi-stage cycling and PT trips.

@@

The mobility survey data {[}3{]} is the basis for this project and
defines the baseline scenario. Despite being conducted in the
pre-pandemic period (2017), this dataset represents the most
comprehensive and up-to-date information on urban mobility in Portuguese
metropolitan areas (Lisbon and Porto).

We used a method for disaggregating the origins and destinations of
trips between the centroids of two districts (same as ``parish'') to
ensure that a district is not solely characterized by a single point of
origin and destination for its trips. Aggregating all trips into
centroids renders the exercise less realistic, as it excludes a
significant portion of short-distance trips, a prevalent characteristic
of active mode travel {[}8{]}. The OD Jittering method breaks down a
single point (i.e., the centroid of an area) into multiple random points
on the existing and neighboring road network, using OpenStreetMap as a
reference. This method then distributes the volume of trips within the
district among the randomly generated origin-destination pairs.

@@

The selected route choice algorithm was the
\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package} {[}9{]},
which allows for great flexibility in configuring estimated route types,
and which proven to provide most accurate route networks for the city of
Lisbon {[}10{]}. \texttt{r5r} can calculate multi-modal routes using PT
combined with other modes. It enables the identification of the most
direct or safest cycling routes, using the Level of Traffic
Stress\footnote{see
  \href{https://docs.conveyal.com/learn-more/traffic-stress}{docs.conveyal.com/learn-more/traffic-stress}.}
(LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest
(e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g.,
routes shared with motorized traffic). The routes were estimated for the
base scenario for both types of networks: \emph{direct} and \emph{safe},
using LTS 4 and LTS 3, respectively. Different routing profiles enable
decision-makers to plan for different bicycle user typologies and/or for
different city cycling maturity levels {[}11{]}.

@@

For the cycling legs of the journey (first and last legs),
socio-environmental impacts were estimated, using the HEAT for Cycling
tool v5.0 {[}12{]} from the World Health Organization, and the
\href{https://github.com/HEAT-WHO/HEAT_heatr_api}{\texttt{HEATaaS} R
package}\footnote{\texttt{HEATaaS} is under development. For more
  information contact
  \href{https://heatwalkingcycling.org}{heatwalkingcycling.org}.}. The
HEAT tool provided estimates on the shifting from car to cycling for a
short term time horizon (i.e., one year) and the long term (i.e., ten
years). We considered two dimensions: \emph{social} --- including the
physical activity of cyclists, air pollution exposure, and road
casualties; \emph{environmental} --- including CO\textsubscript{2}eq
emissions and other pollutants.

@@

To estimate the car emissions, we used the EMEP/EEA's COPERT software v5
methods and reference values {[}13{]} for a Tier 3 detail level. We used
a family-size vehicle, EURO standard, and gasoline or diesel fuel. All
trips were considered to be made under urban conditions and at an
average speed of 15 km/h during rush hour periods. Since the average
distance traveled per trip influences the overconsumption and emissions
from cold-start engine operation, we estimated energy and emission
factors for different ranges of trips at 500-meter intervals. In
addition, we assumed an occupancy rate of 1.6 passengers \emph{per} car
{[}3{]}.

Regarding PT, we considered the emission factor values reported in the
environmental and sustainability reports of the PT operators in the LMA
{[}14--17{]}.

@@

The conversion of avoided emissions into avoided welfare loss and
respective monetary valuation was based on the EU Guide to Cost-benefit
Analysis {[}18{]} and the best up-to-date reference values for the
various gases {[}18--20{]}. The social impacts are in avoided premature
mortality. This result is finally monetized using the \emph{Statistical
Value of Life} for Portugal {[}21{]}. We updated all the monetary
reference values of the literature based on the annual inflation rate in
Portugal for 2022\footnote{See
  \href{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ipc}{Statistics
  Portugal tool for inflation rate estimates between years}.}, and our
10-years estimations assumed a discount rate of 5\% and inflation of
3\%.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-MARTENS2007326}{}}%
\CSLLeftMargin{1. }%
\CSLRightInline{Martens, K.: Promoting bike-and-ride: The dutch
experience. Transportation Research Part A: Policy and Practice. 41,
326--338 (2007).
https://doi.org/\href{https://doi.org/10.1016/j.tra.2006.09.010}{10.1016/j.tra.2006.09.010}.}

\leavevmode\vadjust pre{\hypertarget{ref-RIETVELD200071}{}}%
\CSLLeftMargin{2. }%
\CSLRightInline{Rietveld, P.: The accessibility of railway stations: The
role of the bicycle in the netherlands. Transportation Research Part D:
Transport and Environment. 5, 71--75 (2000).
https://doi.org/\href{https://doi.org/10.1016/S1361-9209(99)00019-X}{10.1016/S1361-9209(99)00019-X}.}

\leavevmode\vadjust pre{\hypertarget{ref-IMOB}{}}%
\CSLLeftMargin{3. }%
\CSLRightInline{INE:
\href{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ine_publicacoes\&PUBLICACOESpub_boui=349495406\&PUBLICACOESmodo=2\&xlang=pt}{Mobilidade
e funcionalidade do território nas {Áreas Metropolitanas do Porto e de
Lisboa}: 2017}. {Instituto National de Estatística}, Lisboa (2018).}

\leavevmode\vadjust pre{\hypertarget{ref-ENMAC}{}}%
\CSLLeftMargin{4. }%
\CSLRightInline{Presidência do Conselho de Ministros: Resolução do
conselho de ministros n.º 131/2019,
\url{https://files.dre.pt/1s/2019/08/14700/0004600081.pdf}, (2019).}

\leavevmode\vadjust pre{\hypertarget{ref-felix2023}{}}%
\CSLLeftMargin{5. }%
\CSLRightInline{Félix, R., Lovelace, R., Moura, F.: {biclaR - Ferramenta
de apoio ao planeamento da rede ciclável na área metropolitana de
Lisboa}, \url{https://biclar.tmlmobilidade.pt}, (2022).}

\leavevmode\vadjust pre{\hypertarget{ref-lovelace2017}{}}%
\CSLLeftMargin{6. }%
\CSLRightInline{Lovelace, R., Goodman, A., Aldred, R., Berkoff, N.,
Abbas, A., Woodcock, J.: The Propensity to Cycle Tool: An open source
online system for sustainable transport planning. Journal of Transport
and Land Use. 10, (2017).
https://doi.org/\href{https://doi.org/10.5198/jtlu.2016.862}{10.5198/jtlu.2016.862}.}

\leavevmode\vadjust pre{\hypertarget{ref-goodman2019}{}}%
\CSLLeftMargin{7. }%
\CSLRightInline{Goodman, A., Rojas, I.F., Woodcock, J., Aldred, R.,
Berkoff, N., Morgan, M., Abbas, A., Lovelace, R.: Scenarios of cycling
to school in england, and associated health and carbon impacts:
Application of the {`}propensity to cycle tool{'}. Journal of Transport
\& Health. 12, 263--278 (2019).
https://doi.org/\href{https://doi.org/10.1016/j.jth.2019.01.008}{10.1016/j.jth.2019.01.008}.}

\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022Jittering}{}}%
\CSLLeftMargin{8. }%
\CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Jittering: A
computationally efficient method for generating realistic route networks
from origin-destination data. Findings. (2022).
https://doi.org/\href{https://doi.org/10.32866/001c.33873}{10.32866/001c.33873}.}

\leavevmode\vadjust pre{\hypertarget{ref-r5r}{}}%
\CSLLeftMargin{9. }%
\CSLRightInline{Pereira, R.H.M., Saraiva, M., Herszenhut, D., Braga,
C.K.V., Conway, M.W.: r5r: Rapid realistic routing on multimodal
transport networks with R5 in r. Findings. (2021).
https://doi.org/\href{https://doi.org/10.32866/001c.21262}{10.32866/001c.21262}.}

\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022exploring}{}}%
\CSLLeftMargin{10. }%
\CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Exploring
jittering and routing options for converting origin-destination data
into route networks: Towards accurate estimates of movement at the
street level. The International Archives of the Photogrammetry, Remote
Sensing and Spatial Information Sciences. XLVIII-4/W1-2022, 279--286
(2022).
https://doi.org/\href{https://doi.org/10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}{10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}.}

\leavevmode\vadjust pre{\hypertarget{ref-felix2017}{}}%
\CSLLeftMargin{11. }%
\CSLRightInline{Félix, R., Moura, F., Clifton, K.J.: Typologies of urban
cyclists: Review of market segmentation methods for planning practice.
Transportation Research Record. 2662, 125--133 (2017).
https://doi.org/\href{https://doi.org/10.3141/2662-14}{10.3141/2662-14}.}

\leavevmode\vadjust pre{\hypertarget{ref-HEAT}{}}%
\CSLLeftMargin{12. }%
\CSLRightInline{Kahlmeier, S., Götschi, T., Cavill, N., Castro
Fernandez, A., Brand, C., Rojas Rueda, D., Woodcock, J., Kelly, P.,
Lieb, C., Oja, P., others:
\href{https://www.euro.who.int/__data/assets/pdf_file/0010/352963/Heat.pdf}{Health
economic assessment tool ({HEAT}) for walking and for cycling: Methods
and user guide on physical activity, air pollution, injuries and carbon
impact assessments}. (2017).}

\leavevmode\vadjust pre{\hypertarget{ref-COPERT}{}}%
\CSLLeftMargin{13. }%
\CSLRightInline{Ntziachristos, L., Samaras, Z.: {EMEP/EEA} air pollutant
emission inventory guidebook 2019,
\url{https://www.emisia.com/utilities/copert/documentation/}, (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-Carris2019s}{}}%
\CSLLeftMargin{14. }%
\CSLRightInline{Carris:
\href{https://www.carris.pt/media/dkdp2wbg/dnf_carris2019_rv5.pdf}{Relatório
de sustentabilidade 2019 - demostração não financeira}. {Carris -
Companhia Carris de Ferro de Lisboa, E.M., S.A.} (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-Metro2019s}{}}%
\CSLLeftMargin{15. }%
\CSLRightInline{Metropolitano de Lisboa:
\href{https://www.metrolisboa.pt/wp-content/uploads/2021/01/relatorio_integrado_2019.pdf}{Relatório
integrado 2019}. {Metropolitano de Lisboa, E.P.E.} (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-CP2019s}{}}%
\CSLLeftMargin{16. }%
\CSLRightInline{CP:
\href{https://www.cp.pt/StaticFiles/Institucional/2_gestao_sustentavel/1_RelatoriosSustentabilidade/relatorio-de-sustentabilidade-2019.pdf}{Relatório
de sustentabilidade 2019}. {CP - Comboios de Portugal, E.P.E.} (2020).}

>>>>

\documentclass[review, doubleblind, 3p,
authoryear]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%

\usepackage[hyphens]{url}

  \journal{Computers, Environment and Urban Studies} % Sets Journal name

\usepackage{graphicx}
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
% TODO: Currently lineno needs to be loaded after amsmath because of conflict
% https://github.com/latex-lineno/lineno/issues/5
\usepackage{lineno} % add
  \linenumbers % turns line numbering on
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={biclaR: Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=blue,
            pdfborder={0 0 0}}

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

@@

\usepackage{subfig}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\biboptions{authoryear}
\usepackage{url}
\usepackage{float}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{makecell}
\usepackage{multirow}

@@

\begin{frontmatter}

  \title{biclaR: Estimating the socio-environmental impacts of car
substitution by bicycle and public transit using open tools}
    \author[CERIS]{Rosa Félix\corref{cor1}%
  \corref{cor1}%
  }
   \ead{rosamfelix@tecnico.ulisboa.pt} 
    \author[CERIS]{Filipe Moura%
  %
  }
   \ead{fmoura@tecnico.ulisboa.pt} 
    \author[ITS]{Robin Lovelace%
  %
  }
   \ead{r.lovelace@uleeds.uk} 
      \affiliation[CERIS]{CERIS, Instituto Superior Técnico -
Universidade de Lisboa, Av Rovisco Pais 1, 1049-001 Lisboa, Portugal}
    \affiliation[ITS]{Institute for Transport Studies, University of
Leeds. 34-40 University Rd, Leeds LS2 9JT, UK}
    \cortext[cor1]{Corresponding author}

  \begin{abstract}
  A high proportion of car trips can be replaced by a combination of
  public transit and cycling for the first-and-last mile. This paper
  estimates the potential for cycling combined with public transit (PT)
  as a substitute for car trips in the Lisbon metropolitan area and
  assesses its socio-environmental impacts using open data and open
  source tools. A decision support tool that facilitates the design and
  development of a metropolitan cycling network was developed
  (\emph{biclaR}). The social and environmental impacts were assessed
  using the \emph{HEAT for Cycling} and the \emph{HEAT as a Service}
  tools. The impacts of shifting car trips to PT were also estimated and
  monetized. The results indicate that 10\% of trips could be made by
  bicycle + PT combination. Shifting to cycling for the first-and-last
  mile stages can reduce annual CO\textsubscript{2}eq emissions from
  6,000 tons/day, with benefits over 10 years of at least €230 million.
  For the PT leg, the transfer from car avoids of up to 20,500 tons of
  CO\textsubscript{2}eq emissions per year. This evidence can support
  policymakers to prioritize interventions that reduce the reliance on
  private motor vehicles.\\
  \end{abstract}
    \begin{keyword}
    Active transport \sep Intermodality \sep First and last
mile \sep Health economic assessment \sep Environmental impacts \sep 
    Open data and methods
  \end{keyword}

 \end{frontmatter}

Please make sure that your manuscript follows the guidelines in the
Guide for Authors of the relevant journal. It is not necessary to
typeset your manuscript in exactly the same way as an article, unless
you are submitting to a camera-ready copy (CRC) journal.

For detailed instructions regarding the elsevier article class, see
\url{https://www.elsevier.com/authors/policies-and-guidelines/latex-instructions}

@@

Combining public transportation (PT) and cycling for the first and last
mile in metropolitan areas can significantly replace private car trips
(Martens, 2007; Rietveld, 2000). This approach requires interventions
and programs to make bicycling more appealing, and the resulting public
investments can have significant social and environmental benefits.

According to the latest mobility survey conducted in 2018 (INE, 2018),
the LMA registered a total of 5.3 million daily trips, with only 0.5\%
by bicycle. Car modal share was 58.4\%, while PT accounted for 15.5\%.
The number of intra-municipal trips --- with origin and destination in
the same municipality --- amounts to 3.5 million trips. This exceeds the
number of inter-municipal trips (1.8 million trips), involving travel
between different municipalities. Cars and public transport are the most
used modes for intercity trips, with cars being the predominant choice
for all journeys.

To achieve the cycling targets set by the Portuguese national cycling
strategy for 2025 and 2030 (4\% and 10\%, respectively) (Presidência do
Conselho de Ministros, 2019), the Lisbon's Metropolitan Department of
Transport introduced \emph{biclaR}\footnote{See
  \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}.}, a
decision support tool that facilitates the design and development of a
metropolitan cycling network (Félix et al., 2022).

\emph{biclaR} builds on the Propensity to Cycle Tool\footnote{See
  \href{https://www.pct.bike/}{pct.bike}.} (PCT), a web application and
research project funded by the UK's Department for Transport in 2015
which launched nationally in 2017 as part of the government's Cycling
and Walking Investment Strategy. The PCT initially used only
origin-destination data for commuting trips as the basis of estimates of
cycling potential at zone, route and route network levels (Lovelace et
al., 2017). The PCT has been extended to include cycling potential for
travel to school in England (Goodman et al., 2019) and other trip types
in other countries.\footnote{See \href{https://www.npt.scot}{npt.scot}
  and \href{https://cruse.bike}{cruse.bike} for examples of the PCT in
  Scotland and Ireland that include estimates of cycling for other
  purposes.} However, to the best of our knowledge, this is the first
time that the method has been integrated with public transport data
using multi-modal routing to estimate the potential and benefits of
multi-stage cycling and PT trips.

@@

The mobility survey data (INE, 2018) is the basis for this project and
defines the baseline scenario. Despite being conducted in the
pre-pandemic period (2017), this dataset represents the most
comprehensive and up-to-date information on urban mobility in Portuguese
metropolitan areas (Lisbon and Porto).

We used a method for disaggregating the origins and destinations of
trips between the centroids of two districts (same as ``parish'') to
ensure that a district is not solely characterized by a single point of
origin and destination for its trips. Aggregating all trips into
centroids renders the exercise less realistic, as it excludes a
significant portion of short-distance trips, a prevalent characteristic
of active mode travel (Lovelace et al., 2022b). The OD Jittering method
breaks down a single point (i.e., the centroid of an area) into multiple
random points on the existing and neighboring road network, using
OpenStreetMap as a reference. This method then distributes the volume of
trips within the district among the randomly generated
origin-destination pairs.

@@

The selected route choice algorithm was the
\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package} (Pereira
et al., 2021), which allows for great flexibility in configuring
estimated route types, and which proven to provide most accurate route
networks for the city of Lisbon (Lovelace et al., 2022a). \texttt{r5r}
can calculate multi-modal routes using PT combined with other modes. It
enables the identification of the most direct or safest cycling routes,
using the Level of Traffic Stress\footnote{see
  \href{https://docs.conveyal.com/learn-more/traffic-stress}{docs.conveyal.com/learn-more/traffic-stress}.}
(LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest
(e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g.,
routes shared with motorized traffic). The routes were estimated for the
base scenario for both types of networks: \emph{direct} and \emph{safe},
using LTS 4 and LTS 3, respectively. Different routing profiles enable
decision-makers to plan for different bicycle user typologies and/or for
different city cycling maturity levels (Félix et al., 2017).

@@

For the cycling legs of the journey (first and last legs),
socio-environmental impacts were estimated, using the HEAT for Cycling
tool v5.0 (Kahlmeier et al., 2017) from the World Health Organization,
and the
\href{https://github.com/HEAT-WHO/HEAT_heatr_api}{\texttt{HEATaaS} R
package}\footnote{\texttt{HEATaaS} is under development. For more
  information contact
  \href{https://heatwalkingcycling.org}{heatwalkingcycling.org}.}. The
HEAT tool provided estimates on the shifting from car to cycling for a
short term time horizon (i.e., one year) and the long term (i.e., ten
years). We considered two dimensions: \emph{social} --- including the
physical activity of cyclists, air pollution exposure, and road
casualties; \emph{environmental} --- including CO\textsubscript{2}eq
emissions and other pollutants.

@@

To estimate the car emissions, we used the EMEP/EEA's COPERT software v5
methods and reference values (Ntziachristos \& Samaras, 2020) for a Tier
3 detail level. We used a family-size vehicle, EURO standard, and
gasoline or diesel fuel. All trips were considered to be made under
urban conditions and at an average speed of 15 km/h during rush hour
periods. Since the average distance traveled per trip influences the
overconsumption and emissions from cold-start engine operation, we
estimated energy and emission factors for different ranges of trips at
500-meter intervals. In addition, we assumed an occupancy rate of 1.6
passengers \emph{per} car (INE, 2018).

Regarding PT, we considered the emission factor values reported in the
environmental and sustainability reports of the PT operators in the LMA
(Carris, 2020; CP, 2020; Metropolitano de Lisboa, 2020; Transtejo,
2014).

@@

The conversion of avoided emissions into avoided welfare loss and
respective monetary valuation was based on the EU Guide to Cost-benefit
Analysis (Sartori et al., 2014) and the best up-to-date reference values
for the various gases (Bickel et al., 2006; Nash et al., 2003; Sartori
et al., 2014). The social impacts are in avoided premature mortality.
This result is finally monetized using the \emph{Statistical Value of
Life} for Portugal (Silva et al., 2021). We updated all the monetary
reference values of the literature based on the annual inflation rate in
Portugal for 2022\footnote{See
  \href{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ipc}{Statistics
  Portugal tool for inflation rate estimates between years}.}, and our
10-years estimations assumed a discount rate of 5\% and inflation of
3\%.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-bickel2006}{}}%
Bickel, P., Friedrich, R., Burgess, A., Fagiani, P., Hunt, A., Jong, G.
D., \& Laird, J. (2006). \emph{{HEATCO - Developing Harmonised European
Approaches for Transport Costing and Project Assessment. Deliverable 5,
Proposal for Harmonised Guidelines.}}
\url{https://trimis.ec.europa.eu/sites/default/files/project/documents/20130122_113653_88902_HEATCO_D5_summary.pdf}

\leavevmode\vadjust pre{\hypertarget{ref-Carris2019s}{}}%
Carris. (2020). \emph{Relatório de sustentabilidade 2019 - demostração
não financeira}. {Carris - Companhia Carris de Ferro de Lisboa, E.M.,
S.A.} \url{https://www.carris.pt/media/dkdp2wbg/dnf_carris2019_rv5.pdf}

\leavevmode\vadjust pre{\hypertarget{ref-CP2019s}{}}%
CP. (2020). \emph{Relatório de sustentabilidade 2019}. {CP - Comboios de
Portugal, E.P.E.}
\url{https://www.cp.pt/StaticFiles/Institucional/2_gestao_sustentavel/1_RelatoriosSustentabilidade/relatorio-de-sustentabilidade-2019.pdf}

\leavevmode\vadjust pre{\hypertarget{ref-felix2023}{}}%
Félix, R., Lovelace, R., \& Moura, F. (2022). \emph{{biclaR - Ferramenta
de apoio ao planeamento da rede ciclável na área metropolitana de
Lisboa}}. {CERIS - Instituto Superior Técnico and Transportes
Metropolitanos de Lisboa}. \url{https://biclar.tmlmobilidade.pt}

\leavevmode\vadjust pre{\hypertarget{ref-felix2017}{}}%
Félix, R., Moura, F., \& Clifton, K. J. (2017). Typologies of urban
cyclists: Review of market segmentation methods for planning practice.
\emph{Transportation Research Record}, \emph{2662}(1), 125--133.
\url{https://doi.org/10.3141/2662-14}

\leavevmode\vadjust pre{\hypertarget{ref-goodman2019}{}}%
Goodman, A., Rojas, I. F., Woodcock, J., Aldred, R., Berkoff, N.,
Morgan, M., Abbas, A., \& Lovelace, R. (2019). Scenarios of cycling to
school in england, and associated health and carbon impacts: Application
of the {`}propensity to cycle tool{'}. \emph{Journal of Transport and
Health}, \emph{12}, 263--278.
\url{https://doi.org/10.1016/j.jth.2019.01.008}

\leavevmode\vadjust pre{\hypertarget{ref-IMOB}{}}%
INE. (2018). \emph{Mobilidade e funcionalidade do território nas {Áreas
Metropolitanas do Porto e de Lisboa}: 2017}. {Instituto National de
Estatística}.
\url{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ine_publicacoes\&PUBLICACOESpub_boui=349495406\&PUBLICACOESmodo=2\&xlang=pt}

\leavevmode\vadjust pre{\hypertarget{ref-HEAT}{}}%
Kahlmeier, S., Götschi, T., Cavill, N., Castro Fernandez, A., Brand, C.,
Rojas Rueda, D., Woodcock, J., Kelly, P., Lieb, C., Oja, P., et al.
(2017). \emph{Health economic assessment tool ({HEAT}) for walking and
for cycling: Methods and user guide on physical activity, air pollution,
injuries and carbon impact assessments}.
\url{https://www.euro.who.int/__data/assets/pdf_file/0010/352963/Heat.pdf}

\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022exploring}{}}%
Lovelace, R., Félix, R., \& Carlino, D. (2022a). Exploring jittering and
routing options for converting origin-destination data into route
networks: Towards accurate estimates of movement at the street level.
\emph{The International Archives of the Photogrammetry, Remote Sensing
and Spatial Information Sciences}, \emph{XLVIII-4/W1-2022}, 279--286.
\url{https://doi.org/10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}

\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022Jittering}{}}%
Lovelace, R., Félix, R., \& Carlino, D. (2022b). Jittering: A
computationally efficient method for generating realistic route networks
from origin-destination data. \emph{Findings}.
\url{https://doi.org/10.32866/001c.33873}

\leavevmode\vadjust pre{\hypertarget{ref-lovelace2017}{}}%
Lovelace, R., Goodman, A., Aldred, R., Berkoff, N., Abbas, A., \&
Woodcock, J. (2017). The Propensity to Cycle Tool: An open source online
system for sustainable transport planning. \emph{Journal of Transport
and Land Use}, \emph{10}(1). \url{https://doi.org/10.5198/jtlu.2016.862}

\leavevmode\vadjust pre{\hypertarget{ref-MARTENS2007326}{}}%
Martens, K. (2007). Promoting bike-and-ride: The dutch experience.
\emph{Transportation Research Part A: Policy and Practice},
\emph{41}(4), 326--338. \url{https://doi.org/10.1016/j.tra.2006.09.010}

\leavevmode\vadjust pre{\hypertarget{ref-Metro2019s}{}}%
Metropolitano de Lisboa. (2020). \emph{Relatório integrado 2019}.
{Metropolitano de Lisboa, E.P.E.}
\url{https://www.metrolisboa.pt/wp-content/uploads/2021/01/relatorio_integrado_2019.pdf}

\leavevmode\vadjust pre{\hypertarget{ref-UNITE}{}}%
Nash, C. et al. (2003). \emph{UNIfication of accounts and marginal costs
for transport efficiency: Final report}. {Institute for Transport
Studies, University of Leeds}.
\url{https://www.its.leeds.ac.uk/projects/unite/downloads/Unite\%20Final\%20Report.pdf}

\leavevmode\vadjust pre{\hypertarget{ref-COPERT}{}}%
Ntziachristos, L., \& Samaras, Z. (2020). \emph{{EMEP/EEA} air pollutant
emission inventory guidebook 2019}. Luxembourg: European Environment
Agency. \url{https://www.emisia.com/utilities/copert/documentation/}

\leavevmode\vadjust pre{\hypertarget{ref-r5r}{}}%
Pereira, R. H. M., Saraiva, M., Herszenhut, D., Braga, C. K. V., \&
Conway, M. W. (2021). r5r: Rapid realistic routing on multimodal
transport networks with R5 in r. \emph{Findings}.
\url{https://doi.org/10.32866/001c.21262}

\leavevmode\vadjust pre{\hypertarget{ref-ENMAC}{}}%
Presidência do Conselho de Ministros. (2019). Resolução do conselho de
ministros n.º 131/2019. In \emph{Diário da República, 1ª série} (Vol.
147, pp. 46--81).
\url{https://files.dre.pt/1s/2019/08/14700/0004600081.pdf}

\leavevmode\vadjust pre{\hypertarget{ref-RIETVELD200071}{}}%
Rietveld, P. (2000). The accessibility of railway stations: The role of
the bicycle in the netherlands. \emph{Transportation Research Part D:
Transport and Environment}, \emph{5}(1), 71--75.
\url{https://doi.org/10.1016/S1361-9209(99)00019-X}
